A unified pseudo-$C_\ell$ framework

### 2018-08-17 doc/Paper/main.tex
addressed Reviewer 2's comments

\lstset{ %
language=python,                % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
numbers=left,                   % where to put the line-numbers
numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=false,           % adds a frame around the code
tabsize=2,          % sets default tabsize to 2 spaces
captionpos=b,           % sets the caption-position to bottom
breaklines=true,        % sets automatic line breaking
breakatwhitespace=false,    % sets if automatic breaks should only happen at whitespace
escapeinside={\%*}{*)}          % if you want to add a comment within your code
}
%\usepackage[colorlinks,linkcolor=red,citecolor=blue,urlcolor=blue ]{hyperref}
 %\hypersetup{draft}
\newcommand{\nv}{\hat{\bf n}}
\newcommand{\wtj}[6]{\left(\begin{array}{ccc} #1 & #2 & #3\\#4 & #5 & #6\end{array} \right)}
\newcommand{\nmt}{{\tt NaMaster}}
\newcommand{\va}{{\mathbf{a}}}
\newcommand{\vb}{{\mathbf{b}}}
\newcommand{\mC}{{\sf C}}
\newcommand{\mM}{{\sf M}}
\newcommand{\mI}{{\mathds{1}}}

@@

    While maximum-likelihood~\citep{1998PhRvD..57.2117B,2003PhRvD..67b3001W} or minimum-variance and quadratic estimators~\citep{1997PhRvD..55.5895T,2000MNRAS.317L..23H,2001PhRvD..64f3001T} can recover the power spectrum with virtually no loss of information, their computational implementation is prohibitively expensive for high resolution data, since these methods scale as $\ell_{\rm max}^6$ and $\ell_{\rm max}^4$ respectively with the maximum multipole $\ell_{\rm max}$. For this reason, pseudo-$C_\ell$ algorithms \citep{1973ApJ...185..413P,2001PhRvD..64h3003W,2001ApJ...561L..11S,2001PhRvD..64h3003W,2002ApJ...567....2H,2002MNRAS.336.1304H,2004MNRAS.350..914C} have become a popular alternative that preserves the $\ell_{\rm max}^3$ scaling of spherical harmonic transforms (and faster in the flat-sky approximation). This method can be used on fields with arbitrary spin \citep{2003ApJS..148..161K} in both curved and flat skies, has been extended to deal with contaminant deprojection in scalar fields \citep{2017MNRAS.465.1847E}, and can be optimized to measure the CMB $E/B$-mode power spectrum \citep{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2009PhRvD..79l3515G,2011PhRvD..83h3003B}. Unfortunately, in spite of the advantages of this method, no public, validated and well-maintained codes exist supporting all of these capabilities. This paper presents a public software package, \nmt, that unifies and extends all of these features providing a general framework to estimate pseudo-$C_\ell$ angular power spectra in a wide variety of scenarios.

    The paper is structured as follows: Section \ref{sec:maths} introduces the pseudo-$C_\ell$ estimator in the curved sky, including mode deprojection and $E/B$ purification. This is then generalized in the flat-sky approximation. In Section \ref{sec:structure} we briefly describe the implementation of these methods on \nmt, as well as its structure and usage. Section \ref{sec:validation} presents the stress tests used to validate the code. We envisage this code to be useful in analysis of both the Cosmic Microwave Background data and in the tomographic large-scale structure from photometric data and this section illustrates not just its validity but also applicability in typical user-case scenarios in both fields. We conclude in Section \ref{sec:discussion}. The appendices provide further details about the mathematical background used throughout the paper, and present some additional validation tests.

@@

    \subsection{Spherical harmonic transform and pseudo-\texorpdfstring{$C_\ell$}{Cl}s}\label{ssec:maths.pcl101}
      Let ${\bf a}(\nv)$ be a spin-$s_a$ quantity defined on the sphere, where $\nv$ is the unit vector pointing in a particular direction $(\theta,\varphi)$. We define its spherical harmonic coefficients through a spherical harmonic transform (SHT) as:
      \begin{align}\label{eq:sht1}
        {\bf a}_{\ell m}&\equiv\mathcal{S}\left[{\bf a}(\nv)\right]^{s_a}_{\ell m}\equiv\int d\nv\,_{s_a}{\sf Y}^{s_a\dag}_{\ell m}(\nv)\,{\bf a}(\nv), \\\label{eq:sht2}
        {\bf a}(\nv)&=\mathcal{S}^{-1}\left[{\bf a}_{\ell m}\right]^{s_a}_{\nv}\equiv\sum_{\ell m}\,{\sf Y}^{s_a}_{\ell m}(\nv)\,{\bf a}_{\ell m}.
      \end{align}
      Note that, in general we will use a vector notation such that, for a complex spin-$s_a$ field $a$, we form the vector ${\bf a}\equiv({\rm Re}(a),{\rm Im}(a))$. The harmonic coefficients above are decomposed in a similar manner into $E$ and $B$ modes: ${\bf a}_{\ell m}\equiv(a^E_{\ell m},a^B_{\ell m})$. The matrices ${\sf Y}^s_{\ell m}$ are defined in terms of the spin-weighed spherical harmonics, and are described in detail in Appendix \ref{app:shts}. To simplify the notation, we will also often abbreviate the pair $(\ell,m)$ as ${\bf l}$.

      For two isotropic fields ${\bf a}$ and ${\bf b}$, the power spectrum is given by their covariance matrix:
      \begin{equation}\label{eq:cldef}
        \left\langle{\bf a}_{\bf l}{\bf b}_{{\bf l}'}^\dag\right\rangle\equiv{\sf C}^{ab}_\ell\delta_{\ell\ell'}\delta_{mm'}.
      \end{equation}
      Note that for a general non-zero spin, vectors $\va$ and $\vb$ have two-components, each of which is a complex number. Nevertheless the rotational invariance requires that elements of the $2\times2$ matrix $\mC$ be real.

      \subsubsection{Power-spectrum estimation}
        The problem we are attempting to solve is how to estimate ${\sf C}^{ab}_\ell$ from a single realization of ${\bf a}$ and ${\bf b}$ measured on a cut sky. In general, the input data for measuring the power spectrum come pixelized on a sphere. Instead of dealing with an infinitely sampled points, we have a finite number of measurements
        \begin{equation}
          \va_i = \va_i^{\rm true} + {\bf n}_i,
        \end{equation}
        where $\va_i^{\rm true}$ is is the true underlying field whose power spectrum we would like to estimate and ${\bf n}_i$ is the noise vector, which in general can be non-white and non-homogeneous. We will denote the pixel covariance of the noise component as ${\sf N}$. Note that for the sake of power-spectrum estimation the there is no conceptual difference between inhomogeneous noise and a survey mask: masked areas can be simply though of as areas of infinite noise. Conversely, when we talk about ``multiplying by the mask'' we really mean a more general process of dividing by the expected variance field.

        The underlying power spectrum in Eq.~\ref{eq:cldef} means that there is a non-trivial true underlying covariance matrix $\mC^{\rm true} = \left<\va^{\rm true} \va^{{\rm true}\dagger}\right>$, which in addition to the noise component leads to the actual data covariance $\mC = \left<\va \va^\dagger\right>=\mC^{\rm true} + {\sf N}$.

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \va^\dagger \mC^{-1}
          \left(\frac{\partial \mC}{\partial B_i} \right)
          \mC^{-1} \va - \rm{Tr}\left[\left(\frac{\partial \mC}{\partial B_i}
          \right) {\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix,the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}
          \mC = \sum_i \left(\frac{\partial \mC}{\partial B_i}\right) + {\sf N},
        \end{equation}
        and the last term correspond to bias of convolved estimates by the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices $\partial {\sf C}/\partial B_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

        The pseudo-$C_\ell$ estimates will be optimal when these steps are close to what an optimal quadratic estimator would do. First, this requires that $\mC$ be close to diagonal, which is true when either the noise is large and uncorrelated, or when the underlying power spectrum is close to white. Multiplication by the mask mixes modes, and replacement of multiplication by the response matrix with band-power averaging, only makes sense, when the mask is ``well-behaved'', i.e. compact with no high-frequency structure in it. If these conditions are not met, the pseudo-$C_\ell$ estimator will still be unbiased (by construction), but it will get progressively less optimal. In practical surveys, however, the loss of optimality rarely exceeds 10-20\% and is typically smaller~\citep{2013MNRAS.435.1857L}.

@@

      \subsubsection{Mode coupling}\label{sssec:maths.pcl101.mcoup}
        Let $v(\nv)$ be a sky mask or weights map for ${\bf a}$, and let us define ${\bf a}^v\equiv v(\nv){\bf a}(\nv)$. As discussed above we start by considering a na\"ive estimator for ${\sf C}^{ab}_\ell$ from the masked fields:
        \begin{equation}\label{eq:pcldef}
          {\rm PCL}_\ell\left({\bf a}^v,{\bf b}^w\right)\equiv\frac{1}{2\ell+1}\sum_{m=-\ell}^\ell{\bf a}^v_{\bf l}{\bf b}^{w\dag}_{\bf l}.
        \end{equation}
        The incomplete sky coverage couples different $\ell$ modes, and makes this estimator biased. The pseudo-$C_\ell$ method is based on computing an analytical prediction for this bias and correcting for it. This is straightforward to do using the results in Appendix \ref{app:shts}. The harmonic coefficients of a masked field are given by
        \begin{equation}
          {\bf a}^v_{\bf l}=\sum_{{\bf l}_1{\bf l}_2}={\sf D}^{s_a}_{{\bf l}{\bf l}_1{\bf l}_2}{\bf a}_{{\bf l}_1}v_{{\bf l}_2},
        \end{equation}
        where ${\sf D}$ is defined in Eq. \ref{eq:dmat}. Using the statistical isotropy of the unmasked field (Eq. \ref{eq:cldef}) and the orthogonality relation Eq. \ref{eq:3jorth}, we obtain:
        \begin{equation}\label{eq:modecoup}
          {\rm vec}\left[\left\langle{\rm PCL}_\ell\left({\bf a}^v_{\bf l},{\bf b}^w_{\bf l}\right)\right\rangle\right]=\sum_{\ell'}{\sf M}^{s_as_b}_{\ell\ell'}\cdot{\rm vec}\left[{\sf C}^{ab}_{\ell'}\right],
        \end{equation}
        where we have defined the vectorization operator \citep{2008PhRvD..77j3013H}
        \begin{equation}
          {\rm vec}\left[\left(\begin{array}{cc}
                          C^{EE} & C^{EB}\\
                          C^{BE} & C^{BB}
                         \end{array}\right)\right]\equiv
          \left(
          \begin{array}{c}
            C^{EE} \\
            C^{EB} \\
            C^{BE} \\
            C^{BB} \\
          \end{array}
          \right).
        \end{equation}

@@

      \subsubsection{Bandpowers}\label{sssec:maths.pcl101.bpws}
        Given the loss of information implicit in masking the originally full-sky field, it is in general not possible to invert Eq. \ref{eq:modecoup} directly. One solution to this problem is to convolve the theory prediction with the mode-coupling matrix at the likelihood level. A more usual approach is instead to bin the coupled pseudo-$C_\ell$ into bandpowers. A bandpower $q$ is defined as a set of $N_q$ multipoles $\vec{\ell}_q\equiv(\ell_q^1,...,\ell_q^{N_q})$ and a set of weights defined at those multipoles $w^\ell_q$, and normalized such that $\sum_{\ell\in\vec{\ell}_q}w_q^\ell=1$. The $q$-th bandpower for the coupled pseudo-$C_\ell$ is then defined as
        \begin{align}\label{eq:bpws}
          {\rm vec}\left[\tilde{\sf B}^{ab}_q\right]&\equiv\sum_{\ell\in\vec{\ell}_q}w_q^\ell\,{\rm vec}\left[{\rm PCL}_{\ell}({\bf a}^v_{\bf l},{\bf b}^w_{\bf l})\right]\\\nonumber
          &=\sum_{\ell\in\vec{\ell}_q}w_q^\ell\sum_{\ell'}{\sf M}^{s_as_b}_{\ell\ell'}\,{\rm vec}\left[{\sf C}^{ab}_{\ell'}\right].
        \end{align}
        One then proceeds by assuming that the true power spectrum is a step-wise function, taking constant values over the multipoles corresponding to each bandpower: ${\sf C}^{ab}_\ell\equiv\sum_q{\sf B}^{ab}_q\Theta(\ell\in\vec{\ell}_q)$ (where $\Theta$ is a binary step function). Inserting this in the equation above, it is then possible to find an unbiased estimator for ${\sf B}^{ab}_q$ in terms of the binned pseudo-bandpowers $\tilde{\sf B}^{ab}_q$:
        \begin{equation}\label{eq:decoup}
          {\rm vec}\left[{\sf B}^{ab}_q\right]=\sum_{q'}\left({\cal M}^{s_as_b}\right)^{-1}_{qq'}\,{\rm vec}\left[\tilde{\sf B}^{ab}_{q'}\right],
        \end{equation}
        where the binned coupling matrix ${\cal M}^{s_as_b}$ is
        \begin{equation}
          {\cal M}^{s_as_b}_{qq'}=\sum_{\ell\in\vec{\ell}_q}\sum_{\ell'\in\vec{\ell}_{q'}}w_q^\ell\,{\sf M}^{s_as_b}_{\ell\ell'}.
        \end{equation}

@@

    \subsection{\texorpdfstring{$E/B$}{E/B} purification}\label{ssec:maths.pureb}
    As described in
    \cite{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2010PhRvD..82b3001Z,2010A&A...519A.104K,2011PhRvD..83h3003B},
    $B$-mode purification refers to the map-level removal of the
    contamination from $E$-modes in a the $B$-mode component of a
    given map caused by an incomplete sky coverage and vice-versa. The
    procedure is particularly useful in situations in which the
    $E$-mode component of the signal is significantly larger than the
    $B$ modes, as is the case in CMB. In this case, removing the
    leakage at the power spectrum level (i.e. the standard
    pseudo-$C_\ell$ approach) produces a sub-optimal estimator in
    which the variance in the $B$-mode power spectrum is dominated by
    the variance of the leaked $E$-modes. This section presents the
    purification method and the associated modifications to the
    pseudo-$C_\ell$ algorithm as described in the references above. In
    this section we describe the algorithm in terms of purifying
    $B$-modes from $E$-mode contamination, but we note that the
    reverse, i.e. purifying $E$-modes from $B$-mode contamination is
    conceptually exactly the same. Our discussion is also
    specific for the spin-2 fields.

@@

      \subsection{Beams and noise}\label{ssec:maths.beam_noise}
        In any practical scenario, the observed sky map will contain a mixture of signal ${\bf s}$ (i.e. the component whose power spectrum we are really interested in) and noise ${\bf n}$ (i.e. a stochastic contaminant that we do not have a template for). Furthermore, the signal component will often be smoothed on the smallest scales by an instrumental beam. We discuss how to deal with both of these complications here.

>>>>

\lstset{ %
language=python,                % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
numbers=left,                   % where to put the line-numbers
numberstyle=\footnotesize,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
frame=false,           % adds a frame around the code
tabsize=2,          % sets default tabsize to 2 spaces
captionpos=b,           % sets the caption-position to bottom
breaklines=true,        % sets automatic line breaking
breakatwhitespace=false,    % sets if automatic breaks should only happen at whitespace
escapeinside={\%*}{*)}          % if you want to add a comment within your code
}
%\usepackage[colorlinks,linkcolor=red,citecolor=blue,urlcolor=blue ]{hyperref}
 %\hypersetup{draft}
\newcommand{\nv}{\hat{\boldsymbol{\theta}}}
\newcommand{\wtj}[6]{\left(\begin{array}{ccc} #1 & #2 & #3\\#4 & #5 & #6\end{array} \right)}
\newcommand{\nmt}{{\tt NaMaster}}
\newcommand{\va}{{\mathbf{a}}}
\newcommand{\vb}{{\mathbf{b}}}
\newcommand{\mM}{{\sf M}}
\newcommand{\mI}{{\mathds{1}}}

@@

    While maximum-likelihood~\citep{1998PhRvD..57.2117B,2003PhRvD..67b3001W} or minimum-variance and quadratic estimators~\citep{1997PhRvD..55.5895T,2000MNRAS.317L..23H,2001PhRvD..64f3001T} can recover the power spectrum with virtually no loss of information, their computational implementation is prohibitively expensive for high resolution data, since these methods scale as $\ell_{\rm max}^6$ and $\ell_{\rm max}^4$ respectively with the maximum multipole $\ell_{\rm max}$. For this reason, pseudo-$C_\ell$ algorithms \citep{1973ApJ...185..413P,2001PhRvD..64h3003W,2001ApJ...561L..11S,2001PhRvD..64h3003W,2002ApJ...567....2H,2002MNRAS.336.1304H,2004MNRAS.350..914C} have become a popular alternative that preserves the $\ell_{\rm max}^3$ scaling of spherical harmonic transforms (and faster in the flat-sky approximation). This method can be used on fields with arbitrary spin \citep{2003ApJS..148..161K} in both curved and flat skies, has been extended to deal with contaminant deprojection in scalar fields \citep{2017MNRAS.465.1847E}, and can be optimized to measure the CMB $E/B$-mode power spectrum \citep{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2009PhRvD..79l3515G,2011PhRvD..83h3003B}. Unfortunately, in spite of the advantages of this method, while a number of power spectrum codes solving some of these problems have been put forward, no public, validated and well-maintained code exists supporting all of these capabilities consistently. This paper presents a public software package, \nmt, that unifies and extends all of these features providing a general framework to estimate pseudo-$C_\ell$ angular power spectra in a wide variety of scenarios.

    The paper is structured as follows: Section \ref{sec:maths} introduces the pseudo-$C_\ell$ estimator in the curved sky, including mode deprojection and $E/B$ purification. This is then generalized in the flat-sky approximation. In Section \ref{sec:structure} we briefly describe the implementation of these methods on \nmt, as well as its structure and usage. Section \ref{sec:validation} presents the stress tests used to validate the code. We envisage this code to be useful in analysis of both the Cosmic Microwave Background data and in the tomographic large-scale structure from photometric data, so this section illustrates not just its validity but also applicability in typical user-case scenarios in both fields. We conclude in Section \ref{sec:discussion}. The appendices provide further details about the mathematical background used throughout the paper, and present some additional validation tests.

@@

    \subsection{Spherical harmonic transform and pseudo-\texorpdfstring{$C_\ell$}{Cl}s}\label{ssec:maths.pcl101}
      Let ${\bf a}(\nv)$ be a spin-$s_a$ quantity defined on the sphere, where $\nv$ is the unit vector pointing in a particular direction $(\theta,\varphi)$. We define its spherical harmonic coefficients through a spherical harmonic transform (SHT) as:
      \begin{align}\label{eq:sht1}
        {\bf a}_{\ell m}&\equiv\mathcal{S}\left[{\bf a}(\nv)\right]^{s_a}_{\ell m}\equiv\int d\nv\,{\sf Y}^{s_a\dag}_{\ell m}(\nv)\,{\bf a}(\nv), \\\label{eq:sht2}
        {\bf a}(\nv)&=\mathcal{S}^{-1}\left[{\bf a}_{\ell m}\right]^{s_a}_{\nv}\equiv\sum_{\ell m}\,{\sf Y}^{s_a}_{\ell m}(\nv)\,{\bf a}_{\ell m}.
      \end{align}
      Note that, in general we will use vector notation such that, for a complex spin-$s_a$ field $a$, we form the vector ${\bf a}\equiv({\rm Re}(a),{\rm Im}(a))$. The harmonic coefficients above are decomposed in a similar manner into $E$ and $B$ modes: ${\bf a}_{\ell m}\equiv(a^E_{\ell m},a^B_{\ell m})$ (e.g. see Appendix \ref{app:shts} and \cite{1997PhRvD..55.1830Z}). The matrices ${\sf Y}^s_{\ell m}$ are defined in terms of the spin-weighed spherical harmonics, and are described in detail in Appendix \ref{app:shts}. To simplify the notation, we will also often abbreviate the pair $(\ell,m)$ as ${\bf l}$.

      For two isotropic fields ${\bf a}$ and ${\bf b}$, the power spectrum is given by their covariance matrix:
      \begin{equation}\label{eq:cldef}
        \left\langle{\bf a}_{\bf l}{\bf b}_{{\bf l}'}^\dag\right\rangle\equiv{\sf C}^{ab}_\ell\delta_{\ell\ell'}\delta_{mm'}.
      \end{equation}
      Note that for a general non-zero spin, vectors $\va$ and $\vb$ have two-components, each of which is a complex number. Nevertheless the rotational invariance requires that elements of the $2\times2$ matrix ${\sf C}_\ell$ be real.

      \subsubsection{Power-spectrum estimation}
        The problem we are attempting to solve is how to estimate ${\sf C}^{ab}_\ell$ from a single realization of ${\bf a}$ and ${\bf b}$ measured on a cut sky. In general, the input data for measuring the power spectrum come pixelized on a sphere. Instead of dealing with an infinitely sampled field, we have a finite number of measurements
        \begin{equation}
          \va_i = \va_i^{\rm true} + {\bf n}_i,
        \end{equation}
        where $\va_i^{\rm true}$ is is the true underlying field whose power spectrum we would like to estimate and ${\bf n}_i$ is the noise vector, which in general can be non-white and non-homogeneous. We will denote the pixel covariance of the noise component as $\tilde{\sf N}$\footnote{Note that the matrices $\tilde{\sf C}$ and $\tilde{\sf N}$ used in this section are the data and noise pixel-pixel covariance matrices, and therefore are different from the power spectrum matrices, ${\sf C}_\ell$ and ${\sf N}_\ell$, used in the rest of the paper (e.g. Eq. \ref{eq:cldef}).}. Note that for the sake of power-spectrum estimation the there is no conceptual difference between inhomogeneous noise and a survey mask: masked areas can be simply though of as areas of infinite noise. Conversely, when we talk about ``multiplying by the mask'' we really mean a more general process of dividing by the expected variance field.

        The underlying power spectrum in Eq.~\ref{eq:cldef} means that there is a non-trivial true underlying covariance matrix $\tilde{\sf C}^{\rm true} = \left<\va^{\rm true} \va^{{\rm true}\dagger}\right>$, which in addition to the noise component leads to the actual data covariance $\tilde{\sf C} = \left<\va \va^\dagger\right>=\tilde{\sf C}^{\rm true} + \tilde{\sf N}$.

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \va^\dagger \tilde{\sf C}^{-1}\,{\sf P}_i\,\tilde{\sf C}^{-1} \va -
          \rm{Tr}\left[{\sf P}_i\,{\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix,the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}\label{eq:bpws_qmv}
          \tilde{\sf C} = \sum_i {\sf P}_i\,B_i + {\sf N},
        \end{equation}
        and the last term in Eq. \ref{eq:bpws_qmv} correspond to bias of convolved estimates due to the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices ${\sf P}_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

        The pseudo-$C_\ell$ estimates will be optimal when these steps are close to what an optimal quadratic estimator would do. First, this requires that $\tilde{\sf C}$ be close to diagonal, which is true when either the noise is large and uncorrelated, or when the underlying power spectrum is close to white. Multiplication by the mask mixes modes, and replacement of multiplication by the response matrix with band-power averaging, makes sense only when the mask is ``well-behaved'', i.e. compact with no high-frequency structure in it. If these conditions are not met, the pseudo-$C_\ell$ estimator will still be unbiased (by construction), but it will get progressively less optimal. In practical surveys, however, the loss of optimality rarely exceeds 10-20\% and is typically smaller~\citep{2013MNRAS.435.1857L}.

@@

      \subsubsection{Mode coupling}\label{sssec:maths.pcl101.mcoup}
        Let $v(\nv)$ be a sky mask or weights map for ${\bf a}$, and let us define ${\bf a}^v\equiv v(\nv){\bf a}(\nv)$. As discussed above we start by considering a na\"ive estimator for ${\sf C}^{ab}_\ell$ from the masked fields:
        \begin{equation}\label{eq:pcldef}
          {\rm PCL}_\ell\left({\bf a}^v,{\bf b}^w\right)\equiv\frac{1}{2\ell+1}\sum_{m=-\ell}^\ell{\bf a}^v_{\bf l}{\bf b}^{w\dag}_{\bf l}.
        \end{equation}
        The incomplete sky coverage couples different $\ell$ modes, and makes this estimator biased. The pseudo-$C_\ell$ method is based on computing an analytical prediction for this bias and correcting for it. This is straightforward to do using the results in Appendix \ref{app:shts}. The harmonic coefficients of a masked field are given by
        \begin{equation}
          {\bf a}^v_{\bf l}=\sum_{{\bf l}_1{\bf l}_2}={\sf D}^{s_a}_{{\bf l}{\bf l}_1{\bf l}_2}{\bf a}_{{\bf l}_1}v_{{\bf l}_2},
        \end{equation}
        where ${\sf D}$ is defined in Eq. \ref{eq:dmat}. Using the statistical isotropy of the unmasked field (Eq. \ref{eq:cldef}) and the orthogonality relation Eq. \ref{eq:3jorth}, we obtain:
        \begin{equation}\label{eq:modecoup}
          {\rm vec}\left[\left\langle{\rm PCL}_\ell\left({\bf a}^v_{\bf l},{\bf b}^w_{\bf l}\right)\right\rangle\right]=\sum_{\ell'}{\sf M}^{s_as_b}_{\ell\ell'}\cdot{\rm vec}\left[{\sf C}^{ab}_{\ell'}\right],
        \end{equation}
        where we have defined the vectorization operator \citep{2008PhRvD..77j3013H}
        \begin{align}
          &{\rm vec}\left[\left(\begin{array}{cc}
                          C^{EE} & C^{EB}\\
                          C^{BE} & C^{BB}
                         \end{array}\right)\right]\equiv
          \left(
          \begin{array}{c}
            C^{EE} \\
            C^{EB} \\
            C^{BE} \\
            C^{BB} \\
          \end{array}
          \right),\\
          &{\rm vec}\left[\left(\begin{array}{c}
                                 C^{TE}\\C^{TB}
                                \end{array}\right)\right]\equiv
            \left(\begin{array}{c}
                    C^{TE}\\C^{TB}
            \end{array}\right),
            \hspace{6pt}
            {\rm vec}\left[C^{TT}\right]={\rm C}^{TT},
        \end{align}
        and $T$ here stands for any spin-$0$ field.

@@

      \subsubsection{Bandpowers}\label{sssec:maths.pcl101.bpws}
        Given the loss of information implicit in masking the originally curved-sky field, it is in general not possible to invert the mode-coupling matrix in Eq. \ref{eq:modecoup} directly. One solution to this problem is to convolve the theory prediction with the mode-coupling matrix at the likelihood level. A more usual approach is instead to bin the coupled pseudo-$C_\ell$ into bandpowers. A bandpower $q$ is defined as a set of $N_q$ multipoles $\vec{\ell}_q\equiv(\ell_q^1,...,\ell_q^{N_q})$ and a set of weights defined at those multipoles $w^\ell_q$, and normalized such that $\sum_{\ell\in\vec{\ell}_q}w_q^\ell=1$. The $q$-th bandpower for the coupled pseudo-$C_\ell$ is then defined as
        \begin{align}\label{eq:bpws}
          {\rm vec}\left[\tilde{\sf B}^{ab}_q\right]&\equiv\sum_{\ell\in\vec{\ell}_q}w_q^\ell\,{\rm vec}\left[{\rm PCL}_{\ell}({\bf a}^v_{\bf l},{\bf b}^w_{\bf l})\right]\\\nonumber
          \left\langle{\rm vec}\left[\tilde{\sf B}^{ab}_q\right]\right\rangle&=\sum_{\ell\in\vec{\ell}_q}w_q^\ell\sum_{\ell'}{\sf M}^{s_as_b}_{\ell\ell'}\,{\rm vec}\left[{\sf C}^{ab}_{\ell'}\right],
        \end{align}
        where in the second line we have taken the expectation value of first one. One then proceeds by assuming that the true power spectrum is a step-wise function, taking constant values over the multipoles corresponding to each bandpower: ${\sf C}^{ab}_\ell\equiv\sum_q{\sf B}^{ab}_q\Theta(\ell\in\vec{\ell}_q)$ (where $\Theta$ is a binary step function). Inserting this in the equation above, it is then possible to find an unbiased estimator for ${\sf B}^{ab}_q$ in terms of the binned pseudo-bandpowers $\tilde{\sf B}^{ab}_q$:
        \begin{figure*}
          \centering
          \includegraphics[width=0.9\textwidth]{maps_contamination}
          \caption{Depiction of the contaminant deprojection process. {\sl Top left:} underlying true signal map (galaxy overdensity used in Section \ref{sssec:validation.suite.lss}). {\sl Top right:} sky contaminant (dust contamination used in Section \ref{sssec:validation.suite.lss}). {\sl Bottom left:} contaminated map obtained by adding the previous two maps. {\sl Bottom right:} cleaned map obtained through Eq. \ref{eq:clean_f}. Since only one contaminant is considered in this case, the cleaning is almost perfect. Residual contamination at the power spectrum level must still be subtracted in order to avoid a biased estimate.}
          \label{fig:lss_cont}
        \end{figure*}
        \begin{equation}\label{eq:decoup}
          {\rm vec}\left[{\sf B}^{ab}_q\right]=\sum_{q'}\left({\cal M}^{s_as_b}\right)^{-1}_{qq'}\,{\rm vec}\left[\tilde{\sf B}^{ab}_{q'}\right],
        \end{equation}
        where the binned coupling matrix ${\cal M}^{s_as_b}$ is
        \begin{equation}
          {\cal M}^{s_as_b}_{qq'}=\sum_{\ell\in\vec{\ell}_q}\sum_{\ell'\in\vec{\ell}_{q'}}w_q^\ell\,{\sf M}^{s_as_b}_{\ell\ell'}.
        \end{equation}

@@

    \subsection{\texorpdfstring{$E/B$}{E/B} purification}\label{ssec:maths.pureb}
      As described in \cite{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2010PhRvD..82b3001Z,2010A&A...519A.104K,2011PhRvD..83h3003B}, $B$-mode purification refers to the map-level removal of the contamination from $E$-modes in a the $B$-mode component of a given map caused by an incomplete sky coverage and vice-versa\footnote{Note that this is a different effect from the leakage between different polarized and unpolarized channels in CMB observations due to instrumental effects (asymmetric beams, pair differencing, polarization angle calibration etc.)}. The procedure is particularly useful in situations in which the $E$-mode component of the signal is significantly larger than the $B$ modes, as is the case in CMB. In this case, removing the leakage at the power spectrum level (i.e. the standard pseudo-$C_\ell$ approach) produces a sub-optimal estimator in which the variance in the $B$-mode power spectrum is dominated by the variance of the leaked $E$-modes. This section presents the purification method and the associated modifications to the pseudo-$C_\ell$ algorithm as described in the references above. In this section we describe the algorithm in terms of purifying $B$-modes from $E$-mode contamination, but we note that the reverse, i.e. purifying $E$-modes from $B$-mode contamination is conceptually exactly the same. Our discussion is also specific for the spin-2 fields.

@@

      \subsection{Beams and noise}\label{ssec:maths.beam_noise}
        In any practical scenario, the observed sky map will contain a mixture of signal ${\bf s}$ (i.e. the component whose power spectrum we are really interested in) and noise ${\bf n}$ (i.e. a stochastic contaminant that we do not have a template for). Furthermore, the signal component will often be smoothed on the smallest scales by an instrumental beam (although this latter effect is less relevant for galaxy surveys). We discuss how to deal with both of these complications here.

        \subsubsection{Beam deconvolution}\label{sssec:maths.beam_noise.beam}
          Including the effect of an spherically symmetric instrumental beam amounts to a simple redefinition of the mode-coupling matrix
          \begin{equation}
            {\sf M}^{s_as_b}_{\ell_1\ell_2}\rightarrow{\sf M}^{s_as_b}_{\ell_1\ell_2}W_{\ell_2}^{ab},
          \end{equation}
          where $W^{ab}_\ell$ is the product of the harmonic transform of the beams for maps $a$ and $b$. Note that since any continuous tangent vector field on the sphere must, at some location, be zero (the so-called hairy ball theorem), it is impossible to have an anisotropic beam that is uniform over the entire sphere.  However, it is possible that the beam is both anisotropic and spatially-varying. In this case, the pseudo-$C_\ell$ power spetrum estimation will produce biased estimates that will be suppressed by an \emph{effective} beam. Correcting for the beam in such cases can also be done analytically, but we leave this for future work.  In practice, if required, the
          suppression factor can be found by applying the PCL estimator to a a set of mock input maps with the same underlying signal but  with and without application of the beam smoothing (to both data maps and contaminant maps) and taking the ratio.
